Model-free and Bayesian Ensembling Model-based Deep Reinforcement Learning for Particle Accelerator Control Demonstrated on the FERMI FEL

### 2025-12-29 paper/main.tex
refactor: Relocate experiment data to a dedicated `data/` directory and add paper assets.

	\begin{abstract}
		Reinforcement learning holds significant potential in accelerator controls. The primary goal of this paper is to demonstrate the operational application of this approach to accelerator physics problems. Despite the success of model-free reinforcement learning in several domains, sample efficiency remains a limiting factor, a challenge that model-based methods may mitigate.
		We compare well-suited purely model-based to model-free reinforcement learning applied to the intensity optimisation on the FERMI FEL system. We find that the model-based approach demonstrates higher representational power and sample-efficiency, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the algorithms were implemented in a way, which presents increased noise robustness as prevalent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	\section{Deep Reinforcement learning}
	Most commonly, two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while MBRL algorithms learn a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, their applications is mainly restricted to simulated environments due to their low sample-efficiency. However, putting constraints on the algorithms, as discussed in \cref{s:Model-free reinforcement learning}, partly cures the sample inefficiency, turning them into practical tools for real-world problems.\\
	On the other hand, MBRL shows much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so-called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	The goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
		p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
		\label{eq:trajectory_distribution}
	\end{align}

@@

	One of the most crucial points is how model uncertainties are taken into account by the MFRL controller. Several different approaches were tried. One way is to take the mean of the models $\mu_m$ and the standard deviation $\sigma_m$ and sample from $\gauss(\mu_m,\sigma_m)$. Another strategy is to randomly select a model to provide at each training step, which demonstrated the best performance the original implementation of the model ensemble \emph{TRPO} (\emph{ME-TRPO}) algorithm and was used in the experiments labelled \emph{ME-TRPO} \cite{Kurutach2018}. A pessimistic selection would only select the model resulting in the lowest predicted reward (as used, e.g. in \cite{kidambi2020morel}) and showed constant but too slow improvement for our online-training.

@@

	\begin{table}[h!]%The best place to locate the table environment is directly after its first reference in text

@@

		\begin{ruledtabular}
			\begin{tabular}{l p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}}

@@

				\textrm{Noise \newline robustness}&

@@

				\textrm{Sample \newline efficiency}\\
				\colrule

@@

@@

		\end{ruledtabular}
	\end{table}

@@

	\begin{table}[!h]%The best place to locate the table environment is directly after its first reference in text

@@

		\begin{ruledtabular}
			\begin{tabular}{l>{\raggedleft} p{1cm}>{\raggedleft}p{1.8cm}>{\raggedleft} p{1.8cm} >{\raggedleft\arraybackslash} p{1.8cm}}
				%\toprule
				&  \centering Data points (counts)&   \centering Episode length (counts)&   \centering Cumulative reward\newline (arb. units)&   \centering Final\\reward \newline (arb. units)\tabularnewline

@@

	The Bayesian community and the RL community is spending more and more attention to
	new approaches to Bayesian inference. In the `anchored ensembling' technique a regularisation
	term is added to the loss function, which returns a point estimate of
	the Bayesian posterior the maximum a posteriori (MAP) parameter estimate. 
	Using an ensemble of networks and adding noise on either the targets of the regularisation term a distribution of MAP solutions  \cite{Gu2007, Chen2011, Bardsley2012, Pearce2018} is obtained, which mimics true posterior.\\

@@

	\Cref{fig:Compare_models_sizes} shows the impact of the number of models onto the performance on the \emph{inverted pendulum}. The maximum cumulative reward averaged over five different runs tested on the real environment during the training is visualized in dependence of the number of data-points. A number of three models (label \emph{Three}) shows a good trade-off between performance and training time. A single network might not converge or too slowly (label \emph{Single}) and a model of ten shown fast and stable performance in this case (label \emph{Ten}).

@@

	Finally, \cref{fig:comparsion_NAF_AE-DYNA} demonstrates the sample-efficiency of the \emph{AE-DYNA-SAC} and the \emph{NAF} algorithm on the noisy \emph{inverted pendulum}. \emph{AE-DYNA-SAC} converges below 2000 data-points (without noise even below 800) and the \emph{NAF2} starts to perform equally 10000 data-points surpassing the -200 reward threshold. One clearly sees the increased sample-efficiency on this problem using the \emph{AE-DYNA} in contrast to the \emph{NAF2}.

>>>>

	\begin{abstract}
		Reinforcement learning holds significant potential in accelerator controls. The primary goal of this paper is to demonstrate the operational application of this approach to accelerator physics problems. Despite the success of model-free reinforcement learning in several domains, sample efficiency remains a limiting factor, a challenge that model-based methods may mitigate.
		We compare well-suited purely model-based to model-free reinforcement learning applied to the intensity optimisation on the FERMI FEL system. We find that the model-based approach demonstrates higher representational power and sample-efficiency, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the algorithms were implemented in a manner that presents increased noise robustness as prevalent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	\section{Deep Reinforcement learning}
	Most commonly, two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while MBRL algorithms learn a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, their application is mainly restricted to simulated environments due to their low sample-efficiency. However, putting constraints on the algorithms, as discussed in \cref{s:Model-free reinforcement learning}, partly cures the sample inefficiency, turning them into practical tools for real-world problems.\\
	On the other hand, MBRL shows much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so-called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	The goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
		p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
		\label{eq:trajectory_distribution}
	\end{align}

@@

	One of the most crucial points is how model uncertainties are taken into account by the MFRL controller. Several different approaches were tried. One way is to take the mean of the models $\mu_m$ and the standard deviation $\sigma_m$ and sample from $\gauss(\mu_m,\sigma_m)$. Another strategy is to randomly select a model to provide at each training step, which demonstrated the best performance in the original implementation of the model ensemble \emph{TRPO} (\emph{ME-TRPO}) algorithm and was used in the experiments labelled \emph{ME-TRPO} \cite{Kurutach2018}. A pessimistic selection would only select the model resulting in the lowest predicted reward (as used, e.g. in \cite{kidambi2020morel}) and showed constant but too slow improvement for our online-training.

@@

	\begin{table*}[t]%The best place to locate the table environment is directly after its first reference in text

@@

			\begin{tabular}{lcccc}
				\hline\hline

@@

				\textrm{Noise robustness}&

@@

				\textrm{Sample efficiency}\\

@@

				\hline\hline

@@

	\end{table*}

@@

	\begin{table*}[t]%The best place to locate the table environment is directly after its first reference in text

@@

			\begin{tabular}{lrrrr}
				\hline\hline
				&  \centering Data points (counts)&   \centering Episode length (counts)&   \centering Cumulative reward (arb. units)&   \centering Final reward (arb. units)\tabularnewline

@@

	The Bayesian and RL communities are paying increasing attention to new approaches for Bayesian inference. In the `anchored ensembling' technique, a regularization term is added to the loss function, which yields a point estimate of the Bayesian posterior, specifically the maximum a posteriori (MAP) parameter estimate. 
	By employing an ensemble of networks and introducing noise to the regularization term's targets, a distribution of MAP solutions \cite{Gu2007, Chen2011, Bardsley2012, Pearce2018} is obtained, which mimics the true posterior.\\

@@

	\Cref{fig:Compare_models_sizes} shows the impact of the number of models onto the performance on the \emph{inverted pendulum}. The maximum cumulative reward averaged over five different runs tested on the real environment during the training is visualized in dependence of the number of data-points. A number of three models (label \emph{Three}) shows a good trade-off between performance and training time. A single network might not converge or too slowly (label \emph{Single}) and a model of ten showed fast and stable performance in this case (label \emph{Ten}).

@@

	Finally, \cref{fig:comparsion_NAF_AE-DYNA} demonstrates the sample-efficiency of the \emph{AE-DYNA-SAC} and the \emph{NAF} algorithm on the noisy \emph{inverted pendulum}. \emph{AE-DYNA-SAC} converges below 2000 data-points (without noise even below 800) and the \emph{NAF2} starts to perform equally after 10000 data-points surpassing the -200 reward threshold. One clearly sees the increased sample-efficiency on this problem using the \emph{AE-DYNA} in contrast to the \emph{NAF2}.
